Specific examples

### 2023-04-24 inst/gitbook/_book/dads_manual.tex
Big rename

{
\setcounter{tocdepth}{1}
\tableofcontents
}
\hypertarget{dads-disparity-and-diversity-simulations.}{%
\chapter{\texorpdfstring{\texttt{dads}: disparity and diversity simulations.}{dads: disparity and diversity simulations.}}\label{dads-disparity-and-diversity-simulations.}}

@@

\hypertarget{what-is-dads}{%
\section{\texorpdfstring{What is \texttt{dads}?}{What is dads?}}\label{what-is-dads}}

@@

\texttt{dads} aims to be a highly modular and friendly version of all these packages: it allows to simulate disparity and diversity jointly with a vast array of options that can be easily modified by users.
For example, you can easily generate any type of process to generate a trait (BM, OU, something else, etc\ldots) in multiple dependent or independent dimensions through \texttt{"traits"} objects.
You can then specify how the traits should affect disparity through \texttt{"modifiers"} objects. And finally you can create events (like mass extinctions) through \texttt{"events"} objects.
These objects and how to modify them will be detailed throughout this manual.
Finally we are putting an emphasise in the development of this package on the speed and reliability of the functions.

@@

Because their is an infinite way you might want to generate disparity and diversity (different traits, different modifiers and different events), the \texttt{dads} package is designed to make all these parts easy to code separately and integrate them easily in the \texttt{dads} core functions.
This allows you to simulate finely tuned multidimensional data for your specific project!

@@

This manual explains how \texttt{dads} work in as much details as possible and aims to give you the keys to design your own diversity and disparity simulations.

@@

\CommentTok{\#\# Installing the latest version directly from GitHub}
\NormalTok{devtools}\OperatorTok{::}\KeywordTok{install\_github}\NormalTok{(}\StringTok{"TGuillerme/dads"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

If you need help with the package, hopefully the following manual will be useful.
However, parts of this package are still in development and some other parts are probably not covered.
Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}) or if you are a GitHub user, directly create an issue on the \href{https://github.com/TGuillerme/dads}{GitHub page}.
Doing so will not only hopefully help you but also other users since it will help improve this manual!

\hypertarget{how-does-dads-work}{%
\section{\texorpdfstring{How does \texttt{dads} work?}{How does dads work?}}\label{how-does-dads-work}}

Basically, the \texttt{dads} function intakes your personalised \texttt{traits}, \texttt{modifiers} and \texttt{events} to generate your disparity and diversity.
You will find more details about how these objects (\texttt{traits}, \texttt{modifiers} and \texttt{events}) work in the rest of the tutorial but here is a graphical representation of how \texttt{dads} work:

\begin{figure}
\centering
\includegraphics{dads_structure.png}
\caption{Schematised summary of the \texttt{dads} package architecture}
\end{figure}

@@

The \texttt{dads} algorithm is based on a modular version of the birth-death process algorithm.
The birth-death model is a continuous Markov process (i.e.~a continuous random process where the state will change each step are linked through time and change according to some random exponential variables) which is well defined mathematically and \href{https://lukejharmon.github.io/pcm/chapter10_birthdeath/\#section-10.2-the-birth-death-model}{commonly used in biology} but also in \href{https://en.wikipedia.org/wiki/Birth\%E2\%80\%93death_process}{many other fields}.

The algorithm used in \texttt{dads} allows modularity of this process and is based on the following steps (the text in \texttt{courier\ font} is for the name of the process in the algorithm):

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{-1}
\tightlist
\item
  \textbf{Starting the process}: this step is actually non modular and creates a random tree with one tip, one node and one branch length connecting both. This step is used to optimise the rest of the algorithm in terms of speed and memory management. The node, tip and branch resulting from this step are discarded at the end of the simulations.
\item
  \textbf{Selecting a lineage} (\texttt{selecting}): this step selects a tip that is currently not extinct. In a standard birth-death process this is just done randomly, however in \texttt{dads} this can be modified based on the birth-death parameters, the currently available lineages and potential trait values. \emph{For example, it is possible to put a higher probability in selecting a lineage that is next to a lineage that recently went extinct and has a positive trait value.}
  Then go to step 2.
\item
  \textbf{Growing the tree} (\texttt{waiting}): this step grows the tree by a certain amount. It does so by adding the same amount of branch length to all the non-extinct lineages. In an exact birth death process, this is done by drawing a random value from an exponential distribution with a rate of \[\text{number of living lineages} \times (\text{speciation} + \text{extinction parameters})\]. In \texttt{dads}, this can be modified based on the birth-death parameters, the currently available lineages and potential trait values. \emph{For example, it is possible to increase branch length by the ratio of living/extinct fossils and a random number drawn from the range of current trait values.}
  If the total tree length is less than the required tree length, go to step 3. Else go to step 6.
\item
  \textbf{Simulating traits} {[}optional{]}(\texttt{traits}): this step allows to simulate a trait value for the selected lineage from step 1. This is typically not part of a standard birth-death process and is handled via the \texttt{traits} option in \texttt{dads} (see the \protect\hyperlink{maketraits}{\texttt{make.traits}} chapter for more details).
  Then go to step 4.
\item
  \textbf{Speciating} (\texttt{speciating}): in this step, the selected lineage from step 1 has the option of speciating or going extinct. In a standard birth-death process, this happens by randomly drawing a value between 0 and 1 as a way to trigger speciation relatively to the birth-death parameters:
\end{enumerate}

@@

Again, in \texttt{dads}, this process can be modified based on the birth-death parameters, the currently available lineages and potential trait values. \emph{For example, the lineage can only speciate if it's trait value is positive, regardless of whether speciation or extinction has been triggered}.
Then go to step 5.
5. If the number of lineages is less than required number of lineages, then go to step 1. Else go to step 6.
6. The simulation has stopped because it reached the required amount of lineages or time.

@@

One of the simplest things to do with the \texttt{dads} package is just to simulate a birth death tree.
For that you can use the function \texttt{dads} and specify your stopping rule.
The stopping rule simply tells the birth death process to step whenever it reaches one of these three conditions:

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Running the birth death simulation}
\NormalTok{my\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ my\_stop\_rule)}
\end{Highlighting}
\end{Shaded}

@@

\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-6-1.pdf}

@@

People familiar with the \href{https://lukejharmon.github.io/pcm/chapter10_birthdeath/}{birth-death models} might have noticed that we did not specify two important things here: the speciation parameter (sometimes called ``lambda'' or ``birth'') and the extinction parameter (sometimes called ``mu'', ``death'' or ``background extinction'').
By default \texttt{dads} runs a pure birth model (the speciation is set to 1 and the extinction to 0).
However, you can easily change that by specifying your new birth death parameters:

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Generating a birth death tree with extinctions:}
\NormalTok{my\_tree \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ my\_params, }\DataTypeTok{stop.rule =}\NormalTok{ my\_stop\_rule)}
\CommentTok{\#\# Visualising the new tree}
\KeywordTok{plot}\NormalTok{(my\_tree)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-8-1.pdf}

@@

Chances are that you want to also simulate traits (disparity) along with your diversity (otherwise, we suggest using the \href{https://github.com/tanja819/TreeSim/}{\texttt{TreeSim}} package that provides many more birth death models).
Simulating traits is not much more complicated in \texttt{dads}: you'll simply need to create a \texttt{"traits"} object using the \texttt{make.traits} function.
These objects can have increasing complexity (see the rest of this tutorial) but we will keep it simple here.

@@

\begin{verbatim}
##  ---- dads traits object ---- 
## 1 trait for 1 process (A) with one starting value (0).
\end{verbatim}

@@

\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-10-1.pdf}

By default, this trait is called ``A''.
This is not a really good name but you'll see more about specifying trait names later on.
If this is what the process should look like (theoretically) you can then add its \texttt{"traits"} object to our previous \texttt{dads} function to generate the tree and the traits:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Simulate disparity and diversity}
\NormalTok{my\_data \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ my\_params,}
                \DataTypeTok{stop.rule =}\NormalTok{ my\_stop\_rule,}
                \DataTypeTok{traits    =}\NormalTok{ my\_trait)}
\end{Highlighting}
\end{Shaded}

@@

\begin{verbatim}
##  ---- dads object ---- 
## Simulated diversity data (x$tree):
## 
## Phylogenetic tree with 20 tips and 19 internal nodes.
## 
## Tip labels:
##   t1, t2, t3, t4, t5, t6, ...
## Node labels:
##   n1, n2, n3, n4, n5, n6, ...
## 
## Rooted; includes branch lengths.
## 
## Simulated disparity data (x$data):
## 1 trait for 1 process (A) with one starting value (0).
\end{verbatim}

@@

\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-12-1.pdf}

@@

\begin{verbatim}
##  ---- dads modifiers object ---- 
## No modifiers applied to the branch length, selection and speciation processes (default).
\end{verbatim}

This will not do much to our simulations compared to the previous trait and tree simulation but we can provide our modifiers object to the \texttt{dads} function:

@@

\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\CommentTok{\#\# Simulate disparity and diversity}
\NormalTok{default\_data \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction\_}\DecValTok{02}\NormalTok{,}
                     \DataTypeTok{stop.rule =}\NormalTok{ living\_}\DecValTok{20}\NormalTok{,}
                     \DataTypeTok{traits    =}\NormalTok{ BM\_trait,}
                     \DataTypeTok{modifiers =}\NormalTok{ default\_modifiers)}
\NormalTok{default\_data}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads object ---- 
## Birth death process with modifiers:
## speciation: 1.
## extinction: 0.2.
## No modifiers applied to the branch length, selection and speciation processes (default).
## 
## Simulated diversity data (x$tree):
## 
## Phylogenetic tree with 24 tips and 23 internal nodes.
## 
## Tip labels:
##   t1, t2, t3, t4, t5, t6, ...
## Node labels:
##   n1, n2, n3, n4, n5, n6, ...
## 
## Rooted; includes branch lengths.
## 
## Simulated disparity data (x$data):
## 1 trait for 1 process (A) with one starting value (0).
\end{verbatim}

@@

First we need to create a condition function to trigger this (non) speciation event.
We can do that by specifying our \texttt{condition} function (when the ancestor is trait is negative) apply our modification (here the \texttt{staying.alive} function).
For that we can use the \texttt{parent.traits} utility function that is optimised for accessing traits in the birth death process (but you can of course write your own).
This function intakes the \texttt{trait.values} and \texttt{parent.lineage} arguments, two arguments that you can leave named as they are to facilitate \texttt{dads}' understanding of what you want to assess:

@@

We can then provide these two functions (the condition \texttt{negative.ancestor} and how to modify the speciation event when this condition is met \texttt{staying.alive}).
If you are an advances \texttt{dads} user, you can design your own \texttt{speciation} function but if you just want to use a normal \texttt{speciation} function, you can use the default one from \texttt{dads} called\ldots{} \texttt{speciation}.

@@

\begin{verbatim}
##  ---- dads modifiers object ---- 
## Default branch length process.
## Default selection process.
## Speciation process is set to speciation with a condition (negative.ancestor) and a modifier (going.extinct).
\end{verbatim}

Note that the \texttt{make.modifiers} function tests whether the input is compatible with \texttt{dads} by default so unless you have an error message, your \texttt{modifiers} will work!
We can now simulate our tree and traits with our modifier: species will go extinct if their ancestor have a negative trait value:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\CommentTok{\#\# Simulate disparity and diversity}
\NormalTok{biased\_data \textless{}{-}}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction\_}\DecValTok{02}\NormalTok{,}
                    \DataTypeTok{stop.rule =}\NormalTok{ living\_}\DecValTok{20}\NormalTok{,}
                    \DataTypeTok{traits    =}\NormalTok{ BM\_trait,}
                    \DataTypeTok{modifiers =}\NormalTok{ negatives\_extinct)}
\NormalTok{biased\_data}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads object ---- 
## Birth death process with modifiers:
## speciation: 1.
## extinction: 0.2.
## Default branch length process.
## Default selection process.
## Speciation process is set to speciation with a condition (negative.ancestor) and a modifier (going.extinct).
## 
## Simulated diversity data (x$tree):
## 
## Phylogenetic tree with 22 tips and 21 internal nodes.
## 
## Tip labels:
##   t1, t2, t3, t4, t5, t6, ...
## Node labels:
##   n1, n2, n3, n4, n5, n6, ...
## 
## Rooted; includes branch lengths.
## 
## Simulated disparity data (x$data):
## 1 trait for 1 process (A) with one starting value (0).
\end{verbatim}

@@

\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-20-1.pdf}

@@

In \texttt{dads}, traits are simulating by providing a \texttt{traits} object to the \texttt{traits} argument.
This object is generated using the \texttt{make.traits}.

\begin{quote}
\textbf{What is a trait in biology?} This is one of these epistemological question like what \emph{is} a species that is not covered by all means in this package.
For the purpose of \emph{this manual}, \textbf{biological trait} can be any coherent process of any number of dimensions which can or cannot be independent.
Basically anything can be a trait as well as it's defined by the user (e.g.~one trait can be a 20 dimensional Brownian process that's correlated to a 1D OU process - which is another trait).
For the purpose of the syntax in \emph{this package} though, a \textbf{trait in \texttt{dads}} is a column in the output trait matrix.
For example, one might be interested in a biological trait that is ``beak shape'' which is defined as one complex biological trait that is the Procrustes superimposition of 4 3D landmarks and 300 3D semi-landmarks.
This single ``beak shape'' \textbf{biological trait} is then composed of 912 \textbf{traits in \texttt{dads}} (i.e.~a matrix of \((4+300)\times3\) columns).
\end{quote}

@@

Trait processes in \texttt{dads} are functions that must always intake the following arguments by default.

@@

\begin{quote}
Note that in this function definition the arguments \texttt{x0} and \texttt{edge.length} have a default value set to \texttt{0} and \texttt{1} respectively. In practice, these arguments are effectively set to the correct values in the \texttt{dads} internal function (i.e.~whatever \texttt{x0} and \texttt{edge.length} are at that specific time of the process) but providing a default can help speed up the algorithms (specifically all the internal checks).
\end{quote}

@@

\hypertarget{using-a-process-in-dads}{%
\subsection{\texorpdfstring{Using a \texttt{"process"} in \texttt{dads}}{Using a "process" in dads}}\label{using-a-process-in-dads}}

You can design your own process as a function (as long as it has a valid syntax).
Alternatively, the \texttt{dads} package proposes inbuilt processes, namely a multidimensional Brownian motion (\texttt{BM.process}) or a a multidimensional Ornstein-Uhlenbeck process (\texttt{OU.process}).
You can find the list of implemented process by looking at the \texttt{?trait.process} manual page in \texttt{R}.

@@

This creates \texttt{"dads"} \texttt{"traits"} objects that you can print, and visualise using the \texttt{plot} function:

@@

\begin{verbatim}
## [1] "dads"   "traits"
\end{verbatim}

@@

\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-25-1.pdf}

Note that you can see the multiple options for plotting the trait process by looking at \texttt{?plot.dads} manual. Furthermore, you can look at what's actually in the object using this specific syntax (this applies to every object handled by the \texttt{dads} package):

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# What\textquotesingle{}s actually in that object?}
\KeywordTok{print.dads}\NormalTok{(my\_trait\_object, }\DataTypeTok{all =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## $main
## $main$A
## $main$A$process
## function (x0 = 0, edge.length = 1, Sigma = diag(length(x0)), 
##     ...) 
## {
##     return(t(MASS::mvrnorm(n = 1, mu = x0, Sigma = sqrt(Sigma^2 * 
##         edge.length), ...)))
## }
## <bytecode: 0x55e278dc9cd8>
## <environment: namespace:dads>
## 
## $main$A$start
## [1] 0
## 
## $main$A$trait_id
## [1] 1
## 
## 
## 
## $background
## NULL
\end{verbatim}

@@

\begin{verbatim}
##  ---- dads traits object ---- 
## 4 traits for 4 processes (A, B, C, D) with one starting value (0).
\end{verbatim}

You can visualise them individually using the \texttt{trait} argument in \texttt{plot.dads}:

@@

\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-28-1.pdf}

>>>>

{
\setcounter{tocdepth}{1}
\tableofcontents
}
\hypertarget{treats-disparity-and-diversity-simulations.}{%
\chapter{\texorpdfstring{\texttt{treats}: disparity and diversity simulations.}{treats: disparity and diversity simulations.}}\label{treats-disparity-and-diversity-simulations.}}

@@

\hypertarget{what-is-treats}{%
\section{\texorpdfstring{What is \texttt{treats}?}{What is treats?}}\label{what-is-treats}}

@@

\texttt{treats} aims to be a highly modular and friendly version of all these packages: it allows to simulate disparity and diversity jointly with a vast array of options that can be easily modified by users.
For example, you can easily generate any type of process to generate a trait (BM, OU, something else, etc\ldots) in multiple dependent or independent dimensions through \texttt{"traits"} objects.
You can then specify how the traits should affect disparity through \texttt{"modifiers"} objects. And finally you can create events (like mass extinctions) through \texttt{"events"} objects.
These objects and how to modify them will be detailed throughout this manual.
Finally we are putting an emphasise in the development of this package on the speed and reliability of the functions.

@@

Because their is an infinite way you might want to generate disparity and diversity (different traits, different modifiers and different events), the \texttt{treats} package is designed to make all these parts easy to code separately and integrate them easily in the \texttt{treats} core functions.
This allows you to simulate finely tuned multidimensional data for your specific project!

@@

This manual explains how \texttt{treats} work in as much details as possible and aims to give you the keys to design your own diversity and disparity simulations.

@@

\CommentTok{\#\# Installing the latest version directly from GitHub}
\NormalTok{devtools}\OperatorTok{::}\KeywordTok{install\_github}\NormalTok{(}\StringTok{"TGuillerme/treats"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

If you need help with the package, hopefully the following manual will be useful.
However, parts of this package are still in development and some other parts are probably not covered.
Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}) or if you are a GitHub user, directly create an issue on the \href{https://github.com/TGuillerme/treats}{GitHub page}.
Doing so will not only hopefully help you but also other users since it will help improve this manual!

\hypertarget{how-does-treats-work}{%
\section{\texorpdfstring{How does \texttt{treats} work?}{How does treats work?}}\label{how-does-treats-work}}

Basically, the \texttt{treats} function intakes your personalised \texttt{traits}, \texttt{modifiers} and \texttt{events} to generate your disparity and diversity.
You will find more details about how these objects (\texttt{traits}, \texttt{modifiers} and \texttt{events}) work in the rest of the tutorial but here is a graphical representation of how \texttt{treats} work:

\begin{figure}
\centering
\includegraphics{treats_structure.png}
\caption{Schematised summary of the \texttt{treats} package architecture}
\end{figure}

@@

The \texttt{treats} algorithm is based on a modular version of the birth-death process algorithm.
The birth-death model is a continuous Markov process (i.e.~a continuous random process where the state will change each step are linked through time and change according to some random exponential variables) which is well defined mathematically and \href{https://lukejharmon.github.io/pcm/chapter10_birthdeath/\#section-10.2-the-birth-death-model}{commonly used in biology} but also in \href{https://en.wikipedia.org/wiki/Birth\%E2\%80\%93death_process}{many other fields}.

The algorithm used in \texttt{treats} allows modularity of this process and is based on the following steps (the text in \texttt{courier\ font} is for the name of the process in the algorithm):

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{-1}
\tightlist
\item
  \textbf{Starting the process}: this step is actually non modular and creates a random tree with one tip, one node and one branch length connecting both. This step is used to optimise the rest of the algorithm in terms of speed and memory management. The node, tip and branch resulting from this step are discarded at the end of the simulations.
\item
  \textbf{Selecting a lineage} (\texttt{selecting}): this step selects a tip that is currently not extinct. In a standard birth-death process this is just done randomly, however in \texttt{treats} this can be modified based on the birth-death parameters, the currently available lineages and potential trait values. \emph{For example, it is possible to put a higher probability in selecting a lineage that is next to a lineage that recently went extinct and has a positive trait value.}
  Then go to step 2.
\item
  \textbf{Growing the tree} (\texttt{waiting}): this step grows the tree by a certain amount. It does so by adding the same amount of branch length to all the non-extinct lineages. In an exact birth death process, this is done by drawing a random value from an exponential distribution with a rate of \[\text{number of living lineages} \times (\text{speciation} + \text{extinction parameters})\]. In \texttt{treats}, this can be modified based on the birth-death parameters, the currently available lineages and potential trait values. \emph{For example, it is possible to increase branch length by the ratio of living/extinct fossils and a random number drawn from the range of current trait values.}
  If the total tree length is less than the required tree length, go to step 3. Else go to step 6.
\item
  \textbf{Simulating traits} {[}optional{]}(\texttt{traits}): this step allows to simulate a trait value for the selected lineage from step 1. This is typically not part of a standard birth-death process and is handled via the \texttt{traits} option in \texttt{treats} (see the \protect\hyperlink{maketraits}{\texttt{make.traits}} chapter for more details).
  Then go to step 4.
\item
  \textbf{Speciating} (\texttt{speciating}): in this step, the selected lineage from step 1 has the option of speciating or going extinct. In a standard birth-death process, this happens by randomly drawing a value between 0 and 1 as a way to trigger speciation relatively to the birth-death parameters:
\end{enumerate}

@@

Again, in \texttt{treats}, this process can be modified based on the birth-death parameters, the currently available lineages and potential trait values. \emph{For example, the lineage can only speciate if it's trait value is positive, regardless of whether speciation or extinction has been triggered}.
Then go to step 5.
5. If the number of lineages is less than required number of lineages, then go to step 1. Else go to step 6.
6. The simulation has stopped because it reached the required amount of lineages or time.

@@

One of the simplest things to do with the \texttt{treats} package is just to simulate a birth death tree.
For that you can use the function \texttt{treats} and specify your stopping rule.
The stopping rule simply tells the birth death process to step whenever it reaches one of these three conditions:

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Running the birth death simulation}
\NormalTok{my\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{stop.rule =}\NormalTok{ my\_stop\_rule)}
\end{Highlighting}
\end{Shaded}

@@

\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-6-1.pdf}

@@

People familiar with the \href{https://lukejharmon.github.io/pcm/chapter10_birthdeath/}{birth-death models} might have noticed that we did not specify two important things here: the speciation parameter (sometimes called ``lambda'' or ``birth'') and the extinction parameter (sometimes called ``mu'', ``death'' or ``background extinction'').
By default \texttt{treats} runs a pure birth model (the speciation is set to 1 and the extinction to 0).
However, you can easily change that by specifying your new birth death parameters:

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Generating a birth death tree with extinctions:}
\NormalTok{my\_tree \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ my\_params, }\DataTypeTok{stop.rule =}\NormalTok{ my\_stop\_rule)}
\CommentTok{\#\# Visualising the new tree}
\KeywordTok{plot}\NormalTok{(my\_tree)}
\end{Highlighting}
\end{Shaded}

\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-8-1.pdf}

@@

Chances are that you want to also simulate traits (disparity) along with your diversity (otherwise, we suggest using the \href{https://github.com/tanja819/TreeSim/}{\texttt{TreeSim}} package that provides many more birth death models).
Simulating traits is not much more complicated in \texttt{treats}: you'll simply need to create a \texttt{"traits"} object using the \texttt{make.traits} function.
These objects can have increasing complexity (see the rest of this tutorial) but we will keep it simple here.

@@

\begin{verbatim}
##  ---- treats traits object ---- 
## 1 trait for 1 process (A) with one starting value (0).
\end{verbatim}

@@

\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-10-1.pdf}

By default, this trait is called ``A''.
This is not a really good name but you'll see more about specifying trait names later on.
If this is what the process should look like (theoretically) you can then add its \texttt{"traits"} object to our previous \texttt{treats} function to generate the tree and the traits:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Simulate disparity and diversity}
\NormalTok{my\_data \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ my\_params,}
                \DataTypeTok{stop.rule =}\NormalTok{ my\_stop\_rule,}
                \DataTypeTok{traits    =}\NormalTok{ my\_trait)}
\end{Highlighting}
\end{Shaded}

@@

\begin{verbatim}
##  ---- treats object ---- 
## Simulated diversity data (x$tree):
## 
## Phylogenetic tree with 20 tips and 19 internal nodes.
## 
## Tip labels:
##   t1, t2, t3, t4, t5, t6, ...
## Node labels:
##   n1, n2, n3, n4, n5, n6, ...
## 
## Rooted; includes branch lengths.
## 
## Simulated disparity data (x$data):
## 1 trait for 1 process (A) with one starting value (0).
\end{verbatim}

@@

\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-12-1.pdf}

@@

\begin{verbatim}
##  ---- treats modifiers object ---- 
## No modifiers applied to the branch length, selection and speciation processes (default).
\end{verbatim}

This will not do much to our simulations compared to the previous trait and tree simulation but we can provide our modifiers object to the \texttt{treats} function:

@@

\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\CommentTok{\#\# Simulate disparity and diversity}
\NormalTok{default\_data \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction\_}\DecValTok{02}\NormalTok{,}
                     \DataTypeTok{stop.rule =}\NormalTok{ living\_}\DecValTok{20}\NormalTok{,}
                     \DataTypeTok{traits    =}\NormalTok{ BM\_trait,}
                     \DataTypeTok{modifiers =}\NormalTok{ default\_modifiers)}
\NormalTok{default\_data}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- treats object ---- 
## Birth death process with modifiers:
## speciation: 1.
## extinction: 0.2.
## No modifiers applied to the branch length, selection and speciation processes (default).
## 
## Simulated diversity data (x$tree):
## 
## Phylogenetic tree with 24 tips and 23 internal nodes.
## 
## Tip labels:
##   t1, t2, t3, t4, t5, t6, ...
## Node labels:
##   n1, n2, n3, n4, n5, n6, ...
## 
## Rooted; includes branch lengths.
## 
## Simulated disparity data (x$data):
## 1 trait for 1 process (A) with one starting value (0).
\end{verbatim}

@@

First we need to create a condition function to trigger this (non) speciation event.
We can do that by specifying our \texttt{condition} function (when the ancestor is trait is negative) apply our modification (here the \texttt{staying.alive} function).
For that we can use the \texttt{parent.traits} utility function that is optimised for accessing traits in the birth death process (but you can of course write your own).
This function intakes the \texttt{trait.values} and \texttt{parent.lineage} arguments, two arguments that you can leave named as they are to facilitate \texttt{treats}' understanding of what you want to assess:

@@

We can then provide these two functions (the condition \texttt{negative.ancestor} and how to modify the speciation event when this condition is met \texttt{staying.alive}).
If you are an advances \texttt{treats} user, you can design your own \texttt{speciation} function but if you just want to use a normal \texttt{speciation} function, you can use the default one from \texttt{treats} called\ldots{} \texttt{speciation}.

@@

\begin{verbatim}
##  ---- treats modifiers object ---- 
## Default branch length process.
## Default selection process.
## Speciation process is set to speciation with a condition (negative.ancestor) and a modifier (going.extinct).
\end{verbatim}

Note that the \texttt{make.modifiers} function tests whether the input is compatible with \texttt{treats} by default so unless you have an error message, your \texttt{modifiers} will work!
We can now simulate our tree and traits with our modifier: species will go extinct if their ancestor have a negative trait value:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\CommentTok{\#\# Simulate disparity and diversity}
\NormalTok{biased\_data \textless{}{-}}\StringTok{ }\KeywordTok{treats}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction\_}\DecValTok{02}\NormalTok{,}
                    \DataTypeTok{stop.rule =}\NormalTok{ living\_}\DecValTok{20}\NormalTok{,}
                    \DataTypeTok{traits    =}\NormalTok{ BM\_trait,}
                    \DataTypeTok{modifiers =}\NormalTok{ negatives\_extinct)}
\NormalTok{biased\_data}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- treats object ---- 
## Birth death process with modifiers:
## speciation: 1.
## extinction: 0.2.
## Default branch length process.
## Default selection process.
## Speciation process is set to speciation with a condition (negative.ancestor) and a modifier (going.extinct).
## 
## Simulated diversity data (x$tree):
## 
## Phylogenetic tree with 22 tips and 21 internal nodes.
## 
## Tip labels:
##   t1, t2, t3, t4, t5, t6, ...
## Node labels:
##   n1, n2, n3, n4, n5, n6, ...
## 
## Rooted; includes branch lengths.
## 
## Simulated disparity data (x$data):
## 1 trait for 1 process (A) with one starting value (0).
\end{verbatim}

@@

\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-20-1.pdf}

@@

In \texttt{treats}, traits are simulating by providing a \texttt{traits} object to the \texttt{traits} argument.
This object is generated using the \texttt{make.traits}.

\begin{quote}
\textbf{What is a trait in biology?} This is one of these epistemological question like what \emph{is} a species that is not covered by all means in this package.
For the purpose of \emph{this manual}, \textbf{biological trait} can be any coherent process of any number of dimensions which can or cannot be independent.
Basically anything can be a trait as well as it's defined by the user (e.g.~one trait can be a 20 dimensional Brownian process that's correlated to a 1D OU process - which is another trait).
For the purpose of the syntax in \emph{this package} though, a \textbf{trait in \texttt{treats}} is a column in the output trait matrix.
For example, one might be interested in a biological trait that is ``beak shape'' which is defined as one complex biological trait that is the Procrustes superimposition of 4 3D landmarks and 300 3D semi-landmarks.
This single ``beak shape'' \textbf{biological trait} is then composed of 912 \textbf{traits in \texttt{treats}} (i.e.~a matrix of \((4+300)\times3\) columns).
\end{quote}

@@

Trait processes in \texttt{treats} are functions that must always intake the following arguments by default.

@@

\begin{quote}
Note that in this function definition the arguments \texttt{x0} and \texttt{edge.length} have a default value set to \texttt{0} and \texttt{1} respectively. In practice, these arguments are effectively set to the correct values in the \texttt{treats} internal function (i.e.~whatever \texttt{x0} and \texttt{edge.length} are at that specific time of the process) but providing a default can help speed up the algorithms (specifically all the internal checks).
\end{quote}

@@

\hypertarget{using-a-process-in-treats}{%
\subsection{\texorpdfstring{Using a \texttt{"process"} in \texttt{treats}}{Using a "process" in treats}}\label{using-a-process-in-treats}}

You can design your own process as a function (as long as it has a valid syntax).
Alternatively, the \texttt{treats} package proposes inbuilt processes, namely a multidimensional Brownian motion (\texttt{BM.process}) or a a multidimensional Ornstein-Uhlenbeck process (\texttt{OU.process}).
You can find the list of implemented process by looking at the \texttt{?trait.process} manual page in \texttt{R}.

@@

This creates \texttt{"treats"} \texttt{"traits"} objects that you can print, and visualise using the \texttt{plot} function:

@@

\begin{verbatim}
## [1] "treats"   "traits"
\end{verbatim}

@@

\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-25-1.pdf}

Note that you can see the multiple options for plotting the trait process by looking at \texttt{?plot.treats} manual. Furthermore, you can look at what's actually in the object using this specific syntax (this applies to every object handled by the \texttt{treats} package):

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# What\textquotesingle{}s actually in that object?}
\KeywordTok{print.treats}\NormalTok{(my\_trait\_object, }\DataTypeTok{all =} \OtherTok{TRUE}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## $main
## $main$A
## $main$A$process
## function (x0 = 0, edge.length = 1, Sigma = diag(length(x0)), 
##     ...) 
## {
##     return(t(MASS::mvrnorm(n = 1, mu = x0, Sigma = sqrt(Sigma^2 * 
##         edge.length), ...)))
## }
## <bytecode: 0x55e278dc9cd8>
## <environment: namespace:treats>
## 
## $main$A$start
## [1] 0
## 
## $main$A$trait_id
## [1] 1
## 
## 
## 
## $background
## NULL
\end{verbatim}

@@

\begin{verbatim}
##  ---- treats traits object ---- 
## 4 traits for 4 processes (A, B, C, D) with one starting value (0).
\end{verbatim}

You can visualise them individually using the \texttt{trait} argument in \texttt{plot.treats}:

@@

\includegraphics{treats_manual_files/figure-latex/unnamed-chunk-28-1.pdf}
